---
title: "Exploratory Analysis: StandyBy Data"
author: "Georg Grunsky"
format: pdf
editor: visual
---

## Reading the Data

Two files provided: sickness_table.csv and sickness_table.xlsx Files are apparently the same. Needs to be checked:

Reading the csv-file

```{r}
require(readr)
sickness_csv <- read_csv("C:/Users/ChSnsMgt/iCloudDrive/DLMDWME01/01_data/01_lake/01_use_case_2/sickness_table.csv", 
    col_types = cols(date = col_datetime(format = "%Y-%m-%d")))
head(sickness_csv)
```

Reading the excel-file

```{r}
require(readxl)
sickness_xls <- read_excel("C:/Users/ChSnsMgt/iCloudDrive/DLMDWME01/01_data/01_lake/01_use_case_2/sickness_table.xlsx", 
    col_types = c("numeric", "date", "numeric", 
        "numeric", "numeric", "numeric", "numeric", 
        "numeric"))
head(sickness_xls)
```

Identical check

```{r}
identical(sickness_csv, sickness_xls)
```

Quick identical check says false. Datastructure and Datasummary will be run on both datasets for detailed comparison.

## Datastructure

Print out data structure

```{r}
print("sickness_csv:")
str(sickness_csv)
print("-----------------------------")
print("sickness_xls:")
str(sickness_xls)
```

Writing data dictionary report

```{r}
attributes(sickness_csv)
```

## Data Summary

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
